Adding Sensitivity to 21cm Inteferometric Probes of Reionization by Optimizing Choice of Baselines

### 2017-03-15 Rotation paper/draft_working.tex
save

The earth's rotation causes the baselines to pick up different modes of the sky with time. 
This effect is used extensively to improve uv coverage in imaging with minimum redundancy arrays. In
a maximally redundant array, baselines that are slightly different
rotate into each other at a time delay. Here we present a method to
improve the sensitivity of the maximum redundancy arrays by including
the uv redundancy of different baselines at a time lag.

@@

We can thus rewrite the delayed transformed visibility as 
\small
\[
\boxed{\begin{aligned}V(b,\tau) & =\frac{1}{X^{2}Y}\int_{H}d^{3}rA(\vec{r})\phi(r)I(\vec{r})\exp\left[-2\pi i\left(\frac{b}{c}\cdot\hat{r}+\tau\right)\nu_{r}\right]\end{aligned}
}.
\]
\normalsize
Here $(r_{x},r_{y},r_{z})=(X\hat{s}_{x},X\hat{s}_{y},Y\nu)$, and
$(Xk_{x},Xk_{y},Yk_{z})=\frac{2\pi}{c}(b_{x},b_{y},\eta)$ relate
the cosmological coordinates $r$ and $k$ to the measured coordinates.
We have written $\nu_{r}$ to remind us that $\nu$ and $r$ are related
by Eq. \eqref{eq:nur}. The beam reception pattern $A$ is dimensionless,
normalized to 1 at its peak (zenith), and we assume it to be the same
for all baselines.

@@

\begin{equation}
\boxed{\begin{aligned} & \langle V^{*}(b,\tau)V(b',\tau')\rangle\\
 & =\left(\frac{2k_{B}}{\lambda^{2}}\right)^{2}P(k_{b,\tau})\int\frac{d\Omega d\nu}{X^{2}Y}|A^{*}(\hat{s},\nu)A(\hat{\Gamma s},\nu)||\phi(\nu)|^{2}\\
 & \qquad \qquad \qquad \qquad \exp\left[-i2\pi\nu\left(\hat{s}\cdot\frac{b}{c}-\hat{\Gamma s}\cdot\frac{b'}{c}\right)\right].\end{aligned}}
\label{eq:final}
\end{equation}
So roughly speaking the cross multiplications of visibilities at a time delay
in uv-space is proportional to the power spectrum times the Fourier
transform of the cross multiplied beam pattern. We can then in principle
combine information from different baseline pairs if we correct for
the phase and normalization. As a check, when applied to equivalent baselines,
$b=b'$, $\hat{s}=\hat{\Gamma s}$, and Eq.(\ref{eq:final}) reduces to Eq.(B9) of \cite{paper32}.

@@

@@

To test the robustness of Eq.(\ref{eq:final}), we need to compare the amplitude and phase of the integral
for a pair of baselines with products of simulated visibilities of those baselines. Without loss of generality, we shall use a single frequency channel of $150$MHz for the following analysis.

@@

\caption{Beam fringe pattern of sep2,0(left), sep2,1 at a time delay (middle),
and their conjugate product. Frequency of $\nu=0.15\text{GHz}$ is
chosen. \label{fig:Beam-fringe-pattern}}
\end{figure}
\end{widetext}

@@

To determine that actual relative contribution to sensitivity of these
baseline pairs, we have to take into account of the multiplicities of
these baselines, or the size of the equivalency classes in a mathematical
language. By these we mean how many physical antenna pairs have the
same length and orientation. Looking at Fig. \ref{fig:sensplot}
we see for example sep1,0 will have higher redundancy than sep2,0,
or sep1,1. The latest release of PAPER-64 data uses the 128-equivalent baselines sep2,1,
sep2,0 and sep2,-1 \cite{Ali2015}, and achieved a $2\sigma$ upper
limit of $22.4mK^{2}$. There, the three sets of equivalent baselines
are only cross multiplied by itself. Assuming that each baseline delivers
the same quality of data (i.e. they have the same height of correlation
peaks, which is in our normalization equal to unity), the relative
contribution to sensitivity can be estimated.

@@

\subsection{Validation with PAPER-128 Data}
Unlike the simulated clean EOR signal, real data are dominated by foreground and thermal noise, neither of which turns out to exhibit the correlation pattern seen in Fig.  \ref{fig:numerics}. Foreground exhibits long-range correlations and would lead to high responses at wide range of time-offsets. Instrumental noise fo different baselines/antenna-pairs, on the other hand, are uncorrelated and thus exhibit responses consistent with zero at all time offsets.

\section{Conclusion}

\bibliographystyle{apj}
%\nocite{*}
\bibliography{draft_working}

>>>>

The earth's rotation causes the baselines to pick up different modes of the sky with time. 
This effect is used extensively to improve uv coverage in imaging with minimum redundancy arrays. In
a maximally redundant array, baselines that are slightly different
rotate into each other at a time delay. Here we present a method to
improve the sensitivity of the maximum redundancy arrays by including
the uv redundancy of different baselines at a time lag. Recently the authors of \cite{wterm} proposed an algorithmic search approach to extract information from all near-equivalent baselines. We here propose a simple and elegant theoretical formalism that in addition allows us to identify good baselines to cross-correlate and predict the weight of sensitivity with speed and accuracy.

@@

We can thus rewrite the delayed transformed visibility as 
\small
\[
\begin{aligned}V(b,\tau) & =\frac{1}{X^{2}Y}\int_{H}d^{3}rA(\vec{r})\phi(r)I(\vec{r})\exp\left[-2\pi i\left(\frac{b}{c}\cdot\hat{r}+\tau\right)\nu_{r}\right]\end{aligned}.
\]
\normalsize
Here $(r_{x},r_{y},r_{z})=(X\hat{s}_{x},X\hat{s}_{y},Y\nu)$, and
$(Xk_{x},Xk_{y},Yk_{z})=\frac{2\pi}{c}(b_{x},b_{y},\eta)$ relate
the cosmological coordinates $r$ and $k$ to the measured coordinates.
We have written $\nu_{r}$ to remind us that $\nu$ and $r$ are related
by Eq. \eqref{eq:nur}. The beam reception pattern $A$ is dimensionless,
normalized to 1 at its peak (zenith), and we assume it to be the same
for all baselines.

@@

\begin{equation}
\begin{aligned} & \langle V^{*}(b,\tau)V(b',\tau')\rangle\\
 & =\left(\frac{2k_{B}}{\lambda^{2}}\right)^{2}P(k_{b,\tau})\int\frac{d\Omega d\nu}{X^{2}Y}|A^{*}(\hat{s},\nu)A(\hat{\Gamma s},\nu)||\phi(\nu)|^{2}\\
 & \qquad \qquad \qquad \qquad \exp\left[-i2\pi\nu\left(\hat{s}\cdot\frac{b}{c}-\hat{\Gamma s}\cdot\frac{b'}{c}\right)\right].\end{aligned}
\label{eq:final}
\end{equation}
So roughly speaking the cross multiplications of visibilities at a time delay
in uv-space is proportional to the power spectrum times the Fourier
transform of the cross multiplied beam pattern. We can then in principle
combine information from different baseline pairs if we correct for
the phase and normalization. As a check, when applied to equivalent baselines,
$b=b'$, $\hat{s}=\hat{\Gamma s}$, and Eq.(\ref{eq:final}) reduces to Eq.(B9) of \cite{paper32}.

@@

\subsection{Rephasing \label{sec:rephs}}
Before we discuss tests, there is one subtlety when applying Eq. \ref{eq:final}. The optimal time offset is given by the $\Gamma$ that maximizes the integral. However, first integrating over the spatial dimension $\Omega$, we see the phase term at peak time-offset is inevitably frequency dependent. This frequency dependence of the phase would lead to destructive interference when we integrate over frequency.

The physical origin of this frequency dependent phase is due to the two visibilities having different phase centers. By default the two visibilities are both phased to zenith at the same time. When they are cross-correlated, they must be rephased to the account for the movement of the zenith. Thus in practice we must rephase the visibilities before delay transform.

In Fig.\ref{fig:freqdiff} we compare two channels 0.16 GHz and 0.17 GHz.  Top panels show the equivalent baseline pairs sep1,0 against itself, and bottom panels show sep1,0 against sep1,1. The left two panels have zero rephasing, and the right two are rephased to a time offset of 0.055 sidereal days, the optimal time offset for the given baseline pair. We see that in bottom left panel, although the magnitude, drawn by the wide curves match up, the phases do not. This means the wider the frequency profile, the more destructive the interference, and lower the sensitivity contribution. In the rephased case, the phases of the near-equivalent case match up. We should use the correponding rephasing for each baseline pair maximum sensitivity.

\begin{figure}[H]
\includegraphics[width=1.1\linewidth]{rephs}
\label{fig:freqdiff}
\caption{Comparisons of the peak phases of two different frequencies. Top panels shows equivalent baselines, bottom shows a pair of near-equivalents. The dot-dashed lines of each color are the imaginary parts, and the solid lines are the real parts. For near-equivalent case, we also draw the magnitude by the broad envelops. }
\end{figure}

@@

To test the robustness of Eq.(\ref{eq:final}), we need to compare the amplitude and phase of the integral
for a pair of baselines with products of simulated visibilities of those baselines. For computational simplicity, we shall use a single frequency channel of $150$MHz for the comparison. We may use this simplification without compromising generality because Eq. (\ref{eq:final}) is valid for a range of frequency channels if an only if it's valid for each single channels, provided that different channels are combined with rephasing as desscribed in Sec. \ref{sec:rephs}.

@@

\caption{Beam fringe pattern of sep2,0(left), sep2,1 at a time delay (middle),
and their conjugate product. Frequency of $\nu=0.15\text{GHz}$ is
chosen and only the real components are shown. \label{fig:Beam-fringe-pattern}}
\end{figure}
\end{widetext}

@@

To determine that actual relative contribution to sensitivity of these
baseline pairs, we have to take into account of the multiplicities of
these baselines, or the size of the equivalency classes in a mathematical
language. By these we mean how many physical antenna pairs have the
same length and orientation. Looking at Fig. \ref{fig:sensplot}
we see for example sep1,0 will have higher redundancy than sep2,0,
or sep1,1. The latest release of PAPER-128 data uses the 128-equivalent baselines sep2,1,
sep2,0 and sep2,-1 \cite{Ali2015}, and achieved a $2\sigma$ upper
limit of $22.4mK^{2}$. There, the three sets of equivalent baselines
are only cross multiplied by itself. Assuming that each baseline delivers
the same quality of data (i.e. they have the same height of correlation
peaks, which is in our normalization equal to unity), the relative
contribution to sensitivity can be estimated.

@@

%\subsection{Notes on Data, Foreground and Noise}
%Unlike the simulated clean EOR signal, real data are dominated by foreground %and thermal noise, neither of which turns out to exhibit the correlation %pattern seen in Fig.  \ref{fig:numerics}. Foreground exhibits long-range %correlations and would lead to high responses at wide range of time-offsets. %Instrumental noise for different baselines/antenna-pairs, on the other hand, are uncorrelated and thus exhibit responses consistent with zero at all time offsets. 
%To illustrate, we use the second observing season of PAPER-128 data, taken from Jan. 21st to March 7th of 2014. To illustrate the effect of noise and foreground, we use the data before fringe-rate and delay filtering.

\section{Conclusion}
Redundant arrays are designed to maximize sensitivity. Current generations of redundant radio arrays, such as those probing the power spectrum of the epoch of reionization could benefit from data analysis techniques that improve the sensitivity. We present an intuitive analysis of cross-multiplying baselines that are close in length and orientation to each other. Given an antenna array configuration, our method quickly identifies the best baseline pairs to cross-multiply and predict the expected sensitivity contribution. With the predicted result one can improve existing power-spectrum pipelines through 3 simple steps. 1). Rephase the visibilities prior to delay transforming by the zenith displacement (or more accurately by the phase predicted by our numerical analysis). 2). Shift the visibilities in time. 3) Cross multiply the visibilities of the two baselines  to form the power-spectrum. 4) Finally combine the different baseline pairs by appropriate inverse-covariance weighting that takes into account the predicted sensitivity contributions of each case.

\appendix
\section{\\Hera Analysis}
\begin{widetext}
\begin{figure}[H]
\includegraphics[scale=0.55]{HeraAntpos}

\caption{Planned Hydrogen Epoch of Reionization Array antenna configurations\label{fig:HeraAntpos}}
\end{figure}
\end{widetext}
\bibliographystyle{apj}
%\nocite{*}
\bibliography{draft_working}
